---
title: "MATH/STAT 338: Probability"
subtitle: "Functions of Random Variables"
author: "Anthony Scotina"
date: 
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["my-theme.css", "my-fonts.css"]
    nature:
      countIncrementalSlides: false
      highlightLines: true
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_accent(base_color = "#5E5E5E") #3E8A83?
options(htmltools.preserve.raw = FALSE)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

```{r, echo = FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE
)
```

```{r, include = FALSE}
library(tidyverse)
library(mosaic)
```

<!--
pagedown::chrome_print("~/Dropbox/Teaching/03-Simmons Courses/MATH228-Introduction to Data Science/Lecture Slides/01-Introduction/01-Introduction.html")
-->

class: center, middle, frame

# Transformations

---

# Motivation

All quantities used to estimate *unknown* **population parameters** are *functions* of a sample of *n* random observations that appear in a sample. 

- After applying a **function** to some random variable $Y$, our goal is to find the distribution of the *transformed* random variable. 

<br>

**Examples**

**Sample Mean**: For a set of *n* random variables, $Y_{1},Y_{2},\dots,Y_{n}$, $$\bar{Y}=\frac{1}{n}\sum_{i=1}^{n}Y_{i}.$$

- In this case, $T=Y_{1}+\cdots+Y_{n}$ is a *random variable*, and $\bar{Y}=T/n$ is a *transformation* of the random variable. 

---

# Motivation

All quantities used to estimate *unknown* **population parameters** are *functions* of a sample of *n* random observations that appear in a sample. 

- After applying a **function** to some random variable $Y$, our goal is to find the distribution of the *transformed* random variable. 

<br>

**Examples**

**Linear Regression**: One of the most common modeling techniques involves modeling some *response variable*, $Y$, as a *linear function* of one or more *explanatory variables*, $X$, plus an *error term*, $\epsilon$: $$Y=aX+b+\epsilon$$

- This is also known as a *location-scale* chance of a random variable, $X$.

---

# Motivation

All quantities used to estimate *unknown* **population parameters** are *functions* of a sample of *n* random observations that appear in a sample. 

- After applying a **function** to some random variable $Y$, our goal is to find the distribution of the *transformed* random variable. 

<br>

**Examples**

**Extreme Values**: In many cases, statisticians are interested in the distribution of *extreme observations*, such as the **minimum** or **maximum**: $$Y_{(1)}=\min(Y_{1},\dots,Y_{n})\quad\text{or}\quad Y_{(n)}=\max(Y_{1},\dots,Y_{n})$$

- The random variables $Y_{(i)}$ are called **order statistics**.

---

# The "iid" Assumption

For most remaining examples (unless stated otherwise), we will assume the following:

1. The *population* is **large** relative to the *sample size*. 

2. Random variables obtained through a random sample, $$Y_{1},Y_{2},\dots,Y_{n},$$ are **independent and identically distributed (iid)**. 

--

The **iid assumption** means that the *joint probability function* for $Y_{1},Y_{2},\dots,Y_{n}$ is $$f(y_{1},y_{2},\dots,y_{n})=f(y_{1})f(y_{2})\cdots f(y_{n}),$$ where each random variable shares a common density function, $f(y)$. 

- Note that this is analogous for *discrete random variables* and PMFs, $p(y)$. 

---

class: center, middle, frame

# The Method of Moment-Generating Functions

---

# Moment-Generating Functions

**Recall**

For a random variable, $Y$, the **moment-generating function (MGF)** of $Y$, $m_{Y}(t)$, is defined to be $$m_{Y}(t)=E(e^{tY})$$

<br>

**Theorem**

If $m(t)$ exists, then for any positive integer $k$, $$\left.\frac{d^{k}m(t)}{dt^{k}}\right]_{t=0}=m^{(k)}(0)=E(Y^{k})$$

---

# MGF Theorems

MGFs have use that extend *far beyond* calculating **moments**! 

- Namely, if two random variables share the *same MGF*, then they share the same *distribution*. 

**Theorem** (Uniqueness Theorem)

Let $m_{X}(t)$ and $m_{Y}(t)$ denote the moment-generating functions of random variables, $X$ and $Y$, respectively. 

- If both moment-generating functions exist and $m_{X}(t)=m_{Y}(t)$ *for all values of* $t$, then $X$ and $Y$ have the same probability distribution. 

--

**Theorem** (MGF of Sum of iid RVs)

Let $Y_{1},Y_{2},\dots,Y_{n}$ be *independent* random variables with MGFs $m_{Y_{1}}(t),\dots,m_{Y_{n}}(t)$, respectively. 

- If $U=Y_{1}+Y_{2}+\cdots+Y_{n}$, then $$m_{U}(t)=m_{Y_{1}}(t)\times m_{Y_{2}}(t)\times\cdots\times m_{Y_{n}}(t).$$

---

# Normal and Chi-Square RVs

**Example**

Let $Z\sim Normal(0, 1)$. Use the method of moment-generating functions to show that $Z^{2}\sim\chi^{2}(df=1)$. 

- *In other words*: Show that the MGF for $Z^{2}$ is the same as the MGF for a $\chi^{2}(1)$ random variable. 

--

**Starting Point**

- $Z\sim Normal(0,1)\implies f(z)=\frac{1}{\sqrt{2\pi}}e^{-z^{2}/2}$

- $m_{Z^{2}}(t)=E(e^{tZ^{2}})=\int_{-\infty}^{\infty}e^{tz^{2}}f(z)\,dz$

**Goal**

Show that $m_{Z^{2}}(t)=(1-2t)^{-1/2}$, the MGF for a $\chi^{2}(1)$ RV. 

---

# Exponential and Gamma RVs

**Practice** 

Let $Y_{i}\sim iid\ Exponential(\beta)$, $i=1,\dots,n$ denote the time between customer arrivals at a checkout counter (at *Piéchart Emporium*...). 

Use the method of moment-generating functions to show that $U=\sum_{i=1}^{n}Y_{i}\sim Gamma(n,\beta)$. 

--

**Solution**

Because the $Y_{i}$ are *iid*, $$m_{U}(t)=m_{\sum Y_{i}}(t)=m_{Y_{1}}(t)\times m_{Y_{2}}(t)\times\cdots\times m_{Y_{n}}(t),$$ where $m_{Y_{i}}(t)=(1-\beta t)^{-1}$. 

Therefore, $m_{U}(t)=(1-\beta t)^{-n}$, which follows the form of a $Gamma(n, \beta)$ MGF. 

---

# Sum of Independent Normal RVs

**Theorem**

Let $Y_{1},Y_{2},\dots,Y_{n}$ be **independent** *normally* distributed random variables with $E(Y_{i})=\mu_{i}$ and $Var(Y_{i})=\sigma^{2}_{i}$ (i.e., *not* necessarily iid). Let $a_{1},a_{2},\dots,a_{n}$ be constants.

If $U=\sum_{i=1}^{n}a_{i}Y_{i}$, then $$U\sim Normal\left(\sum_{i=1}^{n}a_{i}\mu_{i},\sum_{i=1}^{n}a_{i}^{2}\sigma_{i}^{2}\right)$$

--

🚨 What is the distribution of $\bar{Y}=\frac{1}{n}\sum_{i=1}^{n}Y_{i}$ in this setting?

---

class: center, middle, frame

# Order Statistics

---

# Common Statistics

The most common statistics used to summarize numerical data are:

- the **sample mean**, $\bar{Y}=\frac{1}{n}\sum_{i=1}^{n}Y_{i}$ (for *center* of distribution)

- the **sample standard deviation**, $s$ (for *spread* of distribution)

Though we are also interested in using *extreme values* to summarize a distribution, such as...

- the **sample minimum**

- the **sample maximum**

- or other *quantiles* in between

--

To study the distributions of these extreme values, let's take a deeper dive into **order statistics**. 

---

# Order Statistics

For random variables $Y_{1},Y_{2},\dots,Y_{n}$, the **order statistics** are the random variables $Y_{(1)},Y_{(2)},\dots,Y_{(n)}$, where:

- $Y_{(1)}=\min(Y_{1},Y_{2},\dots,Y_{n})$

- $Y_{(2)}=$ the second-smallest of $Y_{1},Y_{2},\dots,Y_{n}$

- $\cdots$

- $Y_{(n-1)}=$ the second-largest of $Y_{1},Y_{2},\dots,Y_{n}$

- $Y_{(n)}=\max(Y_{1},Y_{2},\dots,Y_{n})$

--

🚨 For now, we'll assume that the $Y_{i}$ are *iid* and **continuous** RVs with:

- *Distribution function* $F(y)=P(Y\leq y)$

- *Density function* $f(y)=F'(Y)$

---

# PDF for Maximum

First, let's derive the PDF for $$Y_{(n)}=\max(Y_{1},Y_{2},\dots,Y_{n})$$ by *first* finding the distribution function, $P(Y_{(n)}\leq y)$. 

- Because $Y_{(n)}$ is the **maximum** of $Y_{1},Y_{2},\dots,Y_{n}$, the event $(Y_{(n)}\leq y)$ occurs *if and only if* each of the $(Y_{i}\leq y)$ events occur for $i=1,2,\dots,n$: $$P(Y_{(n)}\leq y)=P(Y_{1}\leq y,Y_{2}\leq y,\dots,Y_{n}\leq y)$$

- How can we use the fact that the $Y_{i}$ are **iid** here? 🤔

--

<br>

It turns out that the PDF for $Y_{(n)}$, $g_{(n)}(y)$, is given by: $$g_{(n)}(y)=n[F(y)]^{n-1}f(y)$$

---

# PDF for Minimum

Derive the PDF for $$Y_{(1)}=\min(Y_{1},Y_{2},\dots,Y_{n})$$ by *first* finding the distribution function, $P(Y_{(1)}\leq y)$. 

- [**Hint**: Use the fact that $P(Y_{(1)}\leq y)=1-P(Y_{(1)}>y)$.]

--

<br>

It turns out that the PDF for $Y_{(1)}$, $g_{(1)}(y)$, is given by: $$g_{(1)}(y)=n[1-F(y)]^{n-1}f(y)$$

---

# Uniform and Beta RVs

Suppose that $Y_{i}\sim iid\ Uniform(0,1)$, $i=1,\dots,n$, each with PDF $$f(y)=1,\quad 0\leq y \leq 1.$$

- Find the PDFs for $Y_{(1)}$ and $Y_{(n)}$. What distribution do these order statistics follow? Is it $Uniform(0,1)$?

--

.pull-left[
```{r, fig.height = 2.25, fig.width = 2.5, dpi = 300, echo = FALSE}
unif_samp = runif(10000)
gf_density( ~ unif_samp) + 
  labs(x = "y", y = "f(y)", 
       title = "Y ~ Uniform(0, 1)") + 
  theme_bw()
```
]

.pull-right[
```{r, fig.height = 2.25, fig.width = 2.5, dpi = 300, echo = FALSE}
unif_max = replicate(10000, {
  unif_samp = runif(10)
  max(unif_samp)
})
gf_density( ~ unif_max) + 
  labs(x = "y", y = "f(y)", 
       title = "Y_(n) ~ Beta(n, 1)") + 
  theme_bw()
```
]

---

# Order Statistics for Exponential RVs

**Example**

Electric components of a certain type have a length of life (in hours) $Y~\sim iid\ Exponential(100)$, with PDF given by $$f(y)=\frac{1}{100}e^{-y/100},\quad y >0.$$

**1.** Suppose that two such components operate independently and in series in certain systems (hence, the system fails when *either* component fails). Find the density function for $X=\min(Y_{1},Y_{2})$, the length of life of the system. 

--

- If $Y\sim Exponential$, then $Y_{(1)}\sim Exponential$. 

--

**2.** Suppose that the components operate in parallel (hence, the system does not fail until *both* components fail). Find the density function for $X=\max(Y_{1},Y_{2})$, the length of life of the system. 

--

- If $Y\sim Exponential$, then $Y_{(n)}$ is *not* exponential. 

---

class: center, middle, frame

# Method of Transformations

---

# Method of Transformations

**Theorem**

Let $Y$ have probability density function $f_{Y}(y)$. If $h(y)$ is either strictly *increasing* or *decreasing* for all $y$, then $U=h(Y)$ has density function $$f_{U}(u)=f_{Y}[h^{-1}(u)]\left|\frac{dh^{-1}(u)}{du}\right|.$$

---

# The Log-Normal Distribution

Let $Y\sim Normal(0,1)$. Find the PDF for $U=e^{Y}$. 

1. Find the inverse function, $y=h^{-1}(u)$. 

2. Evaluate $\left|\frac{dh^{-1}(u)}{du}\right|$. 

3. Find $f_{U}(u)$ using the method of transformations. 

--

<br>

**Solution**

- $U=e^{Y}\implies h^{-1}(u)=\log u$

- $\left|\frac{dh^{-1}(u)}{du}\right|=\frac{1}{u}$

---

# Exponential and Weibull RVs

**Practice** (WMS 6.27)

The $Weibull(\alpha,m)$ density function is given by $$f(x)=\frac{1}{\alpha}mx^{m-1}e^{-x^{m}/\alpha},\quad x>0.$$

- Let $Y\sim Exponential(\beta)$. Show that $W=\sqrt{Y}\sim Weibull(\alpha=\beta, m=2)$. 

--

<br>

**Solution**

- $f_{W}(w)=\frac{2}{\beta}we^{-y^{2}/\beta},\ w > 0$

---

# Uniform and Exponential RVs

Let $Y\sim Uniform(0,1)$. Show that $U=-2\log Y \sim Exponential(2)$.

--

<br>

**Solution**

- $U=-2\log Y\implies h^{-1}(u)=e^{-u/2}$

- $\frac{dh^{-1}(u)}{du}=-\frac{1}{2}e^{-u/2}$

- $f_{U}(u)=\frac{1}{2}e^{-u/2},\ u>-$

---

class: center, middle, frame

# Sampling Distributions

---

# Introduction

Previously, we worked through different methods for finding distributions of **functions of random variables**, $Y_{i}$, $i=1,\dots,n$. 

Now we'll treat the $Y_{1},Y_{2},\dots,Y_{n}$ as variables observed in a **random sample** from a **population** of interest. 

- These variables are **iid** (*independent* and *identically distributed*)

--

**Example**

The **sample mean**, $$\bar{y}=\frac{1}{n}\sum_{i=1}^{n}y_{i},$$ is used to estimate the *population mean*, $\mu$. 

- The *goodness* of this estimate depends on the random variables $Y_{1},Y_{2},\dots,Y_{n}$ and how they impact $\bar{Y}=(1/n)\sum Y_{i}$. 

---

# Statistics

The random variable $\bar{Y}$ is an example of a *statistic*, because it is a function of *only* the random variables $Y_{1},Y_{2},\dots,Y_{n}$ and the sample size, $n$ (a constant). 

> A **statistic** is a function of the observable random variables in a sample and known constants. 

--

**Examples**

- The **sample mean**: $\bar{Y}=(1/n)\sum_{i=1}^{n}Y_{i}$

- The **sample variance**: $\frac{1}{n-1}\sum_{i=1}^{n}(Y_{i}-\bar{Y})^{2}$

- The **range**: $R=Y_{(n)}-Y_{(1)}$

--

Because statistics are *functions* of random variables... 

.center[
**All statistics are random variables**!
]

---

# Sampling Distributions

Because **all statistics are random variables**, all statistics have *probability distributions* that illustrate (among other things) how much they *vary from sample to sample*. 

- These "special" probability distributions are called **sampling distributions**. 

--

**Example**

A balanced die is tossed three times, and the number on the uppermost face is recorded each time. Suppose we are interested in the *average of the numbers shown* in the sample of three tosses, $$\bar{Y}=\frac{Y_{1}+Y_{2}+Y_{3}}{3}$$

> Derive the *sampling distribution* of $\bar{Y}$.

---

# Sampling Distributions

```{r, echo = FALSE, fig.align='center', dpi = 300, out.width = "70%"}
set.seed(338)
dice_rolls = replicate(10000, {
  roll = sample(1:6, size = 3, replace = TRUE)
  mean(roll)
})
gf_histogram(~ dice_rolls, binwidth = 1/3, color = "white") + 
  scale_x_continuous(breaks = scales::pretty_breaks()) + 
  labs(title = "Sampling Distribution", 
       subtitle = "Sample Mean of Three Die Tosses", 
       x = expression(bar(Y)), y = "Frequency (out of 10,000 reps)") +
  theme_minimal() + 
  theme(axis.text.x = element_text(size = 12, face = "bold"), 
        axis.text.y = element_text(size = 12, face = "bold"), 
        axis.title = element_text(size = 12, face = "bold"), 
        plot.title = element_text(size = 12, face = "bold"))
```

---

class: center, middle, frame

# Central Limit Theorem

---

# The Sample Mean

**Previous Result**

Let $Y_{1},Y_{2},\dots,Y_{n}$ be *independent* random variables with $E(Y_{i})=\mu$ and $Var(Y_{i})=\sigma^{2}$. Define the **sample mean** as $$\bar{Y}=\frac{1}{n}\sum_{i=1}^{n}Y_{i}.$$ Then:

- $E(\bar{Y})=\mu$

- $Var(\bar{Y})=\sigma^{2}/n$

--

.center[**But what is the *sampling distribution* of the sample mean?!**]

--

If $Y_{i}\sim iid\ Normal(\mu, \sigma^{2})$, then $\bar{Y}\sim Normal(\mu,\sigma^{2}/n)$. 

.center[**But what if the *Y*<sub>i</sub> are *not* Normal?!**]

---

# Central Limit Theorem

Let $Y_{1},Y_{2},\dots,Y_{n}$ be *independent* and *identically distributed* random variables with $E(Y_{i})=\mu$ and $Var(Y_{i})=\sigma^{2}<\infty$. Then $$Z_{n}=\frac{\bar{Y}-\mu}{\sigma/\sqrt{n}}\to Normal(0,1),\ \text{as}\ n\to\infty,$$ where $\bar{Y}=(1/n)\sum_{i=1}^{n}Y_{i}$. 

- **In other words**: For large $n$, the distribution of $\bar{Y}$ *after standardization* approaches a *standard Normal distribution*. 

--

<br>

🚨🚨 The distribution of the $Y_{i}$ can be **anything**, as long as their mean and variance are *finite*. 

- No matter what, the distribution of the *averages* of the $Y_{i}$ will converge to the standard Normal!

.center[
(Though the "starting distribution" and *sample size* **DO** matter...)
]

---

# Simulating the CLT (Normal)

Suppose $Y_{1},Y_{2}\sim iid\ Normal(100, 225)$. That is:

- $E(Y_{i})=100$

- $Var(Y_{i})=225\implies \sigma = 15$

🤔🤔 What does the *sampling distribution* of $Z_{n}$ look like?

```{r}
# One simulation
set.seed(338) # Use for reproducibility!
n = 2; mu = 100; sigma = 15
Y_samp = rnorm(n, mean = mu, sd = sigma)

Z = (mean(Y_samp) - mu)/(sigma/sqrt(n))
Z
```

---

# Simulating the CLT (Normal)

Suppose $Y_{1},Y_{2}\sim iid\ Normal(100, 225)$. That is:

- $E(Y_{i})=100$

- $Var(Y_{i})=225\implies \sigma = 15$

🤔🤔 What does the *sampling distribution* of $Z_{n}$ look like?

```{r}
# MANY simulations!
set.seed(338) # Use for reproducibility!
n = 2; mu = 100; sigma = 15

Z = replicate(10000, {
  Y_samp = rnorm(n, mean = mu, sd = sigma)
  (mean(Y_samp) - mu)/(sigma/sqrt(n))
})
c(mean(Z), var(Z))
```

---

# Simulating the CLT (Exponential)

Suppose $Y_{1},Y_{2}\sim iid\ Exponential(\beta = 2)$. That is:

- $E(Y_{i})=2$

- $Var(Y_{i})=2^{2}=4\implies \sigma=2$

🤔🤔 What does the *sampling distribution* of $Z_{n}$ look like?

```{r}
# MANY simulations!
set.seed(338) # Use for reproducibility!
n = 2; beta = 2
mu = 2; sigma = 2

Z = replicate(10000, {
  Y_samp = rexp(n, rate = 1/beta) #<<
  (mean(Y_samp) - mu)/(sigma/sqrt(n)) #<<
})
c(mean(Z), var(Z))
```

---

# Simulating the CLT (Exponential)

Suppose $Y_{1},Y_{2},\dots,Y_{100}\sim iid\ Exponential(\beta = 2)$. That is:

- $E(Y_{i})=2$

- $Var(Y_{i})=2^{2}=4\implies \sigma=2$

🤔🤔 What does the *sampling distribution* of $Z_{n}$ look like?

```{r}
# MANY simulations!
set.seed(338) # Use for reproducibility!
n = 100; beta = 2 #<<
mu = 2; sigma = 2 

Z = replicate(10000, {
  Y_samp = rexp(n, rate = 1/beta) 
  (mean(Y_samp) - mu)/(sigma/sqrt(n)) 
})
c(mean(Z), var(Z))
```

---

# A Note on Sample Size

👀 What's the deal with the **exponential distribution**?

- Let's see what this "starting distribution" looks like...

.pull-left[
```{r, dpi = 350}
Y = rexp(10000, rate = 1/2)
hist(Y, main = "Y ~ Exp(2)")
```
]

--

.pull-right[
If the $Y_{i}$ come from a strongly **skewed** or *multimodal* distribution, we might need $n$ to be very large before the sampling distribution approaches *Normal(0, 1)*. 

- But if the starting distribution for $Y_{i}$ is already Normal, the sampling distribution for $Z_{n}$ is *exactly* Normal for all $n$!
]

---

# Simulating the CLT (Binomial)

The CLT also applies when the starting distribution is discrete!

Suppose $Y_{1},Y_{2},\dots,Y_{5}\sim iid\ Binomial(n = 20, p = 0.20)$. That is:

- $E(Y_{i})=np = 4$

- $Var(Y_{i})=np(1-p)=3.2\implies \sigma=1.789$

🤔🤔 What does the *sampling distribution* of $Z_{n}$ look like?

```{r}
set.seed(338) # Use for reproducibility!
n = 5; size = 20; p = 0.2
mu = 4; sigma = 1.789

Z = replicate(10000, {
  Y_samp = rbinom(n, size = size, prob = p) 
  (mean(Y_samp) - mu)/(sigma/sqrt(n)) 
})
c(mean(Z), var(Z))
```

---

# Simulating the CLT (Uniform)

Suppose $Y_{1},Y_{2},\dots,Y_{100}\sim iid\ Uniform(a=0, b=1)$. That is:

- $E(Y_{i})=(a+b)/2=1/2$

- $Var(Y_{i})=(b-a)^2/12=1/12\implies\sigma=1/\sqrt{12}$

🤔🤔 What does the *sampling distribution* of $Z_{n}$ look like?

```{r}
set.seed(338) # Use for reproducibility!
n = 100; a = 0; b = 1
mu = 1/2; sigma = 1/sqrt(12)

Z = replicate(10000, {
  Y_samp = runif(n, min = a, max = b) 
  (mean(Y_samp) - mu)/(sigma/sqrt(n)) 
})
c(mean(Z), var(Z))
```

---

# Practice

(WMS, Example 7.9)

The service times for customers coming through a checkout counter in a retail store are independent random variables with mean 1.5 minutes and variance 1.0. 

> Approximate the probability that 100 customers can be served in less than 2 hours of total service time. 

---

class: center, middle, frame

# Sampling Distributions

## Related to the Normal

---

#  Some thms that we've sort of already seen...

1. Let $Y_{1},Y_{2},\dots,Y_{n}$ be a random sample of size $n$ from a **Normal** distribution with mean $\mu$ and variance $\sigma^{2}$. Then $$\bar{Y}=\frac{1}{n}\sum_{i=1}^{n}Y_{i}$$ is normally distributed with mean $\mu_{\bar{Y}}=\mu$ and $\sigma^{2}_{\bar{Y}}=\sigma^{2}/n$. 

--

2. Let $Y_{1},Y_{2},\dots,Y_{n}$ be a random sample of size $n$ from a **Normal** distribution with mean $\mu$ and variance $\sigma^{2}$. Then $Z_{i}=(Y_{i}-\mu)/\sigma$ are *independent*, **standard Normal** random variables, $i=1,2,\dots,n$, and $$\sum_{i=1}^{n}Z_{i}^{2}=\sum_{i=1}^{n}\left(\frac{Y_{i}-\mu}{\sigma}\right)^{2}$$ has a $\chi^{2}$ distribution with $n$ degrees of freedom. 

---

# The Sample Variance

The $\chi^{2}$ distribution is related to the distribution of the **sample variance**, $$S^{2}=\frac{1}{n-1}\sum_{i=1}^{n}(Y_{i}-\bar{Y})^{2}.$$

- In particular, this distribution is used when making *inference* for the *population standard deviation*, $\sigma^{2}$, based on a random sample $Y_{1},Y_{2},\dots,Y_{n}$ from a Normal population. 

--

**Theorem**: Let $Y_{1},Y_{2},\dots,Y_{n}$ be a random sample of size $n$ from a **Normal** distribution with mean $\mu$ and variance $\sigma^{2}$. Then $$\frac{(n-1)S^{2}}{\sigma^{2}}=\frac{1}{\sigma^{2}}\sum_{i=1}^{n}(Y_{i}-\bar{Y})^{2}$$ has a $\chi^{2}$ distribution with $(n-1)$ degrees of freedom. 

---

# Student's t distribution

When the population standard deviation, $\sigma$, is *unknown*, it can be estimated by $S=\sqrt{S^{2}}$, and the quantity $$T=\frac{\bar{Y}-\mu}{S/\sqrt{n}}$$ is used in certain procedures for inference about $\mu$. 

- This quantity, $T$, has a **t distribution** with $n-1$ degrees of freedom!

--

**Definition**: Let $Z$ be a **standard Normal** random variable, and let $W$ be a $\chi^{2}$-distributed random variable with $\nu$ degrees of freedom. Then, if $Z$ and $W$ are *independent*, $$T=\frac{Z}{\sqrt{W/\nu}}$$ is said to have a **t distribution** with $\nu$ degrees of freedom. 

